The cost of school consolidations: impacts of reduced accessibility

### 2024-04-10 paper/Manuscript.qmd
Fixed cross-references before sharing with Chris and Moataz

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: pdf
editor: source
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University
    footnote: "Corresponding Author"
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University
address:
  - code: McMaster University
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
    | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

@@

```{r augmenting-catchment}
ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
              summarize(OTGC = sum(OTGC2011, na.rm = T)), 
            by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))

@@

```{r importing-taz-and-associated-flows}
#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp")) %>% st_transform(crs = 32617)

@@

#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp")) %>%
  st_transform(crs = 32617)

@@

q_3 <- data.frame(
  label = paste0("Q3:\n", c(q_motorized_2011[3], q_nonmotorized_2011[3]|> round(),
                            q_motorized_2016[3], q_nonmotorized_2016[3]|> round())),
  mode   = c("motorized", "non-motorized", "motorized", "non-motorized"),
  year = c("2011", "2011", "2016", "2016"),
  x     = c(q_motorized_2011[3], q_nonmotorized_2011[3] |> round(),
            q_motorized_2016[3], q_nonmotorized_2016[3]|> round()),
  y     = c(0.11, 0.11, 0.11, 0.11)
)
```
```{r tts-theoretical-curve-creation}
x <- data.frame(t = seq(0, 45, 0.1))
X_shorter <-data.frame(t = seq(0, 27, 0.1))

@@

```{r catchments-2011-2016, fig.cap="\\label{fig:Fig1}The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016. OTGC is presented by quartiles. ", out.width="100%", echo=FALSE}

@@

```{r elem-res-rate-2016, fig.cap="\\label{fig:Fig2}The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016. The proportion of low-income households with children under 18 years old is also shown (bottom row). All scales are represented in quartiles.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
```

@@

```{r imped-20112016, fig.cap="\\label{fig:Fig3}The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries. ", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
```
Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.  For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022]. The walk travel times are filtered to only include travel times under 27 minutes. These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.

```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
```

@@

Accessibility is defined as "access to opportunities" or the "potential for spatial interaction". It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
```{=tex}
\begin{equation}
\label{eq:multi-modal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}
```
\noindent where:

@@

```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
```

@@

```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
```

@@

```{r}
SAvail_im_COMM_changes |> st_drop_geometry() |> 
  dplyr::select(c("COMMUNITY",
                  "prev_LIMAT_2016", #prevalance of LIM after-tax for households with children under 18
                  "mode", 
                  "kid_proportion_2016", #the proportion of children 5-14 age, sums to 100%
                  "c_kid_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "V_im_proportion_2016",  #the proportion of V_im, sums to 100%
                  "c_V_im_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "v_im_2016", #the proportion of V_im (in community) /kid_population (in community) = v_im, sums to 100%
                  "c_v_im_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "total_tt_2016", #summed travel time (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_total_tt_perc",  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "emissions_perkid_2016", #summed motorized travel time * 0.1 g/min (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_emissions_perkid_perc"  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  ))
```

>>>>

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: pdf
editor: source
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University SEES
    footnote: "Corresponding Author"
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: University of Toronto
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliation: McMaster University CE
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University SEES
address:
  - code: McMaster University SEES
    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: McMaster University CE
    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: University of Toronto
    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
abstract: | 
  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"),
                          quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),
                                quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

@@

```{r augmenting-catchment, include=FALSE}
ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
              summarize(OTGC = sum(OTGC2011, na.rm = T)), 
            by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))

@@

```{r importing-taz-and-associated-flows, include=FALSE}
#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp"),
                   quiet = TRUE) %>% 
  st_transform(crs = 32617)

@@

#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp"),
                  quiet = TRUE) %>%
  st_transform(crs = 32617)

@@

q_3 <- data.frame(
  label = paste0("Q3:\n", c(q_motorized_2011[3], q_nonmotorized_2011[3]|> round(),
                            q_motorized_2016[3], q_nonmotorized_2016[3]|> round())),
  mode   = c("motorized", "non-motorized", "motorized", "non-motorized"),
  year = c("2011", "2011", "2016", "2016"),
  x     = c(q_motorized_2011[3], q_nonmotorized_2011[3] |> round(),
            q_motorized_2016[3], q_nonmotorized_2016[3]|> round()),
  y     = c(0.11, 0.11, 0.11, 0.11)
)
```

```{r tts-theoretical-curve-creation}
x <- data.frame(t = seq(0, 45, 0.1))
X_shorter <-data.frame(t = seq(0, 27, 0.1))

@@

```{r}
#| label: fig-Fig1
#| fig-cap: "The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016. OTGC is presented by quartiles."
#| echo: false

@@

```{r}
#| label: fig-Fig2
#| fig-cap: "The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016. The proportion of low-income households with children under 18 years old is also shown (bottom row). All scales are represented in quartiles."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
```

@@

```{r}
#| label: fig-Fig3
#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
```
Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.  For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022]. The walk travel times are filtered to only include travel times under 27 minutes. These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.

```{r}
#| label: fig-Fig4
#| fig-cap: "Empirical and theoretical motorized and non-motorized impedance functions."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
```

@@

Accessibility is defined as "access to opportunities" or the "potential for spatial interaction". It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
```{=tex}
\begin{equation}
\label{eq:multi-modal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}
```
\noindent where:

@@

```{r}
#| label: fig-Fig6
#| fig-cap: "Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
```

@@

```{r}
#| label: fig-Fig7
#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
```

@@

```{r include=FALSE}
SAvail_im_COMM_changes |> st_drop_geometry() |> 
  dplyr::select(c("COMMUNITY",
                  "prev_LIMAT_2016", #prevalance of LIM after-tax for households with children under 18
                  "mode", 
                  "kid_proportion_2016", #the proportion of children 5-14 age, sums to 100%
                  "c_kid_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "V_im_proportion_2016",  #the proportion of V_im, sums to 100%
                  "c_V_im_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "v_im_2016", #the proportion of V_im (in community) /kid_population (in community) = v_im, sums to 100%
                  "c_v_im_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "total_tt_2016", #summed travel time (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_total_tt_perc",  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "emissions_perkid_2016", #summed motorized travel time * 0.1 g/min (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_emissions_perkid_perc"  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
  ))
```
